---
title: "Building a Machine Learning Model for Human Activity Recognition (HAR) for Wearable Devices"
author: "Mahfuzul Haque"
date: "23 December 2015"
output: html_document
---

##Introdution

This report explains the process used to build the machine learning model based on the provided training dataset and to predict the unseen cases in the testing dataset. Both the training and testing datasets are obtainined from [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). The data is related to various human activities captured using wearable devices such as _Jawbone Up_, _Nike FuelBand_, _Fitbit_ etc. The training data contains five different activities labelled as _A_, _B_, _C_, _D_, and _E_. The goal of the project was to build a model based on this data and then predicting the activity labels in the testing data. 


```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r, echo=FALSE}
setwd("C:/Users/mahfuzul.haque/Google Drive/KB/Data Science/Code/Practical-Machine-Learning-Project/ds-pml-project")
```


```{r, message=FALSE, warning=FALSE, echo=FALSE}
#source("Submit.R")
library(caret)
library(randomForest)
```

## Getting the data

For this project the training and testing datasets were provided in separate files. We load both the datasets in separate data frames.  

```{r, cache=TRUE}
trainingPath = "pml-training.csv" 
trainingData <- read.csv(trainingPath, na.strings = c("NA", "#DIV/0!", ""), stringsAsFactors=FALSE)

testingPath = "pml-testing.csv" 
testingData <- read.csv(testingPath, na.strings = c("NA", "#DIV/0!", ""), stringsAsFactors=FALSE)
```


## Feature selection 
The labelled training dataset contains 19622 observations and 160 features/variables including the activity label variable. Some of the features are irrelevant to the characteristics of any activity and a large number of features do not contain any value for any observations. Therefore, these variables were removed to improve the efficiency of model training. As shown below, the refined training dataset contains only 53 features including the activity label _classe_.   

```{r, cache=TRUE}
dim(trainingData)

features <- names(trainingData)

selectedFeatures <- c()

for (i in 1:length(features) ) 
  if ( sum(is.na(trainingData[,c(features[i])])) == 0 ) 
    selectedFeatures[length(selectedFeatures) + 1] = features[i]   

selectedFeatures <- selectedFeatures[-c(1,2,3,4,5,6,7)] 

print(selectedFeatures)

refinedTrainingData <- trainingData[, c(selectedFeatures)]

refinedTrainingData$classe <- as.factor(refinedTrainingData$classe)

dim(refinedTrainingData)
```

## Building the model 
To build the model the labelled training dataset was partitioned into two groups. The larger group with 75% of the data was used to train a random forest model with default parameters. The training accuracy of the model was more that 99%. Then the second group of labelled data, not used during model training, was used to measure the accuracy of the model which was also found to be more than 99% as well.      

```{r, cache=TRUE}
set.seed(3433)
inTrain = createDataPartition(refinedTrainingData$classe, p = 3/4)[[1]]

training = refinedTrainingData[inTrain,]
testing = refinedTrainingData[-inTrain,]

dim(training)
dim(testing)

set.seed(62433)

model_rf <- randomForest(classe ~ ., data = training)
print(model_rf)

pred <- predict(model_rf, newdata = testing)

accuracy_rf = sum(pred == testing$classe) / length(pred)

print(accuracy_rf)

```

## Predicting the unlabelled testing data
Finally, the trained model was used to predict the 20 unlabelled cases. As this dataset contains all the 160 features, at first, only the features required by the model were extracted and a refined data frame was constructed. Then the labels for this test data were predicted and written to individual files for submission using the _pml_write_files_ function.    

```{r, echo=FALSE}
# # "B" "A" "B" "A" "A" "E" "D" "B" "A" "A" "B" "C" "B" "A" "E" "E" "A" "B" "B" "B"
```

```{r, cache=TRUE}
dim(testingData)

selectedFeatures <- selectedFeatures[-length(selectedFeatures)]

refinedTestingData <- testingData[, c(selectedFeatures)]

pred2 <- predict(model_rf, newdata = refinedTestingData)

res <- as.character(pred2)
print(res) 

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(res)

```

